On Weighted Multivariate Sign Functions

### 2020-09-20 EJSSubmission/R1/WeightedSign_EJS_Introduction_09_17_20_1.tex
almost submitted R&R

Specifically, we propose using product of the generalized sign function and a weight function derived as a transformation of a data-depth function \citep{ref:DIMACS061_Serfling, ref:AoS00461_ZuoSerfling}. Like data-depth functions, the weight functions used in this paper are non-negative reals defined over $\cX \times \cF$, where $\cF$ is a fixed family of probability measures. For every choice of parameters $\mu \in \cX$ and $\BF \in \cF$, in this paper
%
\begin{align}
R (X_{i}; \mu, \BF) = \mu +  S( X_{i}, \mu) W( X_{i}, \BF).
\label{eqn:rank}
\end{align}
%
is used as a robust surrogate for observation $X_{i}$.
Notice that for the trivial choice $W( x, \BF) = | x - \mu_{\BF} |$, 
 $\mu = \mu_{\BF}$,  we get $R (X_{i}; \mu, \BF) = X_{i}$, the original observations. 
 With the other trivial choice of  $W( x, \BF) \equiv 1$, we get the 
 generalized  sign $R (X_{i}; \mu, \BF) = S(X_{i}, \mu) = S_{i}$. 
 However, in this paper we illustrate how using other weight functions
 can lead to interesting robustness and efficiency trade-offs in a variety of situations. 
 The various technical conditions and assumptions that we impose on the 
 weight function $W( x, \BF)$ are valid for weights derived 
 from three well-known data depth functions: the \textit{half-space depth}, 
 the \textit{Mahalanobis depth}, and the \textit{projection depth}.

@@

We primarily focus on the task of \textit{robust dispersion/scatter estimation} and 
\textit{robust principal component analysis} in this paper. Figure~\ref{fig:Fig1} presents an illustrative example of bivariate data with outliers in the top left panel, where the outliers are marked with red points. In the 
other panels, the generalized sign values of the same data are presented as black points 
on the unit circle, with the outliers again marked with red points. Notice that the 
black points from either the top right or bottom panels have very similar eigenvector 
structure as the original data without the outliers. The green and blue triangles 
are examples of 
the proposed \textit{weighted sign} values: the top right (respectively, bottom left, 
bottom right) panels depict these values where the weights have been generated using 
Mahalnobis depth (respectively Tukey's half-space depth and the projection depth). The 
blue triangles are the weighted sign values of the outliers. Notice that the eigenvectors 
from the weighted signs also capture the pattern from the original data without the 
outliers.

There are two unknown quantities in the generalized sign function as defined in \eqref{eq:GSign}: $\mu$ and $\BF$, To estimate dispersion and its eigen-structure robustly, we must start with a robust estimator for $\mu$. In Section~\ref{Sec:WSQuantiles} we present the case for \textit{weighted spatial quantiles}, which can be defined and studied in very general spaces 
$\cX$. One special case of this is the {\it weighted spatial median}. As a location estimator, it has several interesting robustness properties and can be shown to be more efficient that some existing robust location estimators, thus making it a perfect candidate to estimate $\mu$. On the other hand, setting $\BF$ as $\BF_X$, i.e. the distribution of $X$ has the clear interpretation of differentially weighting observations based on their inlyingness to the overall data distribution. This mirrors the context of how data depth has been used in the literature \cite{LiuPareliusSingh99, ref:DIMACS061_Serfling}. Due to this reason, we assume $\BF \equiv \BF_X$ for the rest of the paper. As and when required, we shall use sample versions of these quantities, stating the theoretical conditions for the corresponding approximation results to hold.

>>>>

Specifically, we propose using product of the generalized sign function and a weight function derived as a transformation of a data-depth function \citep{ref:DIMACS061_Serfling, ref:AoS00461_ZuoSerfling}. Like data-depth functions, the weight functions used in this paper are non-negative reals defined over $\cX \times \cF$, where $\cF$ is a fixed family of probability measures. For every choice of parameters $\mu \in \cX$ and $\BF \in \cF$, in this paper
%
\begin{align}
R (X_{i}; \mu, \BF) = \mu +  S( X_{i}, \mu) W( X_{i}, \BF).
\label{eqn:rank}
\end{align}
%
is used as a robust surrogate for observation $X_{i}$.
Notice that for the trivial choice $W( x, \BF) = | x - \mu_{\BF} |$, 
 $\mu = \mu_{\BF}$,  we get $R (X_{i}; \mu, \BF) = X_{i}$, the original observations. 
 With the other trivial choice of  $W( x, \BF) \equiv 1$, we get the 
 generalized  sign $R (X_{i}; \mu, \BF) = S(X_{i}, \mu) = S_{i}$. 
 However, in this paper we illustrate how using other weight functions
 can lead to interesting robustness and efficiency trade-offs in a variety of situations. 
 The various technical conditions and assumptions that we impose on the 
 weight function $W( x, \BF)$ are valid for weights derived 
 from three well-known data depth functions: the \textit{half-space depth}, 
 the \textit{Mahalanobis depth}, and the \textit{projection depth}. Note that for i.i.d. data, there are three parameters involved here: the location parameter $\mu$ in 
 \eqref{eq:GQSign}, the distribution $\BF$ used for the weight function, and the distribution $\BF_{X}$ of $X_{1}$. For clarity and to mirror the contexts of how data depth has been used in the literature \cite{LiuPareliusSingh99, ref:DIMACS061_Serfling}, we fix $\BF = \BF_{X}$ for this paper, although in Section~\ref{Sec:WSQuantiles} we briefly remark on the case when these  distributions are different. Also note that 
 in many problems of interest, $\BF$ is 
unknown and  data-depths are computed using $\BF_{n}$, however, under very standard regularity conditions (for example, see assumptions \textbf{B1-B3} below), the 
properties of $W (x, \BF)$ and $W (x, \BF_{n})$ are close enough for both asymptotic theory and practical applicability.

@@

We primarily focus on the task of \textit{robust dispersion/scatter estimation} and 
\textit{robust principal component analysis} in this paper. Figure~\ref{fig:Fig1} presents an illustrative example of bivariate data with outliers in the top left panel, where the outliers are marked with red points. In the 
other panels, the generalized sign values of the same data are presented as black points 
on the unit circle, with the outliers again marked with red points. Notice that the 
black points from either the top right or bottom panels have very similar eigenvector 
structure as the original data without the outliers. The green and blue triangles 
are examples of 
the proposed \textit{weighted sign} values: the top right (respectively, bottom left, 
bottom right) panels depict these values where the weights have been generated using 
Mahalanobis depth (respectively Tukey's half-space depth and the projection depth). The 
blue triangles are the weighted sign values of the outliers. Notice that the eigenvectors 
from the weighted signs also capture the pattern from the original data without the 
outliers.

There are two unknown quantities in the generalized sign function as defined in \eqref{eq:GSign}: $\mu$ and $\BF$, To estimate dispersion and its eigen-structure robustly, we must start with a robust estimator for $\mu$. In Section~\ref{Sec:WSQuantiles} we present the case for \textit{weighted spatial quantiles}, which can be defined and studied in very general spaces 
$\cX$. One special case of this is the {\it weighted spatial median}. As a location estimator, it has several interesting robustness properties and can be shown to be more efficient that some existing robust location estimators, thus making it a perfect candidate to estimate $\mu$. 
%On the other hand, setting $\BF$ as $\BF_X$, i.e. the distribution of $X$ has the clear interpretation of differentially weighting observations based on their inlyingness to the overall data distribution. This mirrors the context of how data depth has been used in the literature \cite{LiuPareliusSingh99, ref:DIMACS061_Serfling}. Due to this reason, we assume $\BF \equiv \BF_X$ for the rest of the paper. As and when required, we shall use sample versions of these quantities, stating the theoretical conditions for the corresponding approximation results to hold.

### 2020-09-20 EJSSubmission/R1/WeightedSign_EJS_WeightedSpatialQuantiles_09_17_20_1.tex
almost submitted R&R

Suppose the open unit sphere in $\cX$ is given by 
$\textrm{int} {\cX}_{0; 1} = \{ x \in \cX: | x | < 1 \}$, and let
 $u \in \textrm{int} {\cX}_{0; 1}$. We also fix the set of probability measures $\cM$, 
 and select $\BF \in \cM$. Consider a random element $X \in \cX$. 
We define the $(u, \BF)$-th \textit{weighted spatial quantile} of $\cX$ 
as the minimizer $q (u, \BF) \in \cX$ of 
\baq
\Psi (q; u, \BF) = \BE \Bigl[ W (X, \BF) \bigl\{ | X - q | 
+ \langle u, X - q \rangle \bigr\} \Bigr].
\label{eq:WeightedQuantile}
\eaq
%
%\rework{
\noindent
This is a natural generalization of the spatial median \cite{ref:JASA96862_Chaudhuri, ref:Biometrika48414_Haldane,ref:AoS97435_Koltchinskii} ($W(X, \BF) \equiv 1$ and $u = {\bf 0}_p$), or more general spatial quantiles \cite{ref:AoS17591_Cardotetal_Median_HilbertSpace, ref:AoS141203_ChakrabortyChaudhuri_Banach_Quantile, ref:Bernoulli152308_Minsker_Median_Banach} ($W(X, \BF) \equiv 1$).

@@

\item[(A3)] $\BE W^{2} (X, \BF)  S (X; q) S^{T} (X; q)$ exists for all $q$ in a neighborhood of $q_{0}$. 
\end{description}
 These assumptions are very broad-based and general. The first one essentially requires 
 the existence of a population parameter, the second one requires that the minimization 
 approach is meaningful in the population, and the third one essentially requires that the 
 weight function has a finite variance. No further restrictions are placed on the tuning parameter $\BF$ or the choice of the weight function.

 \begin{Theorem}\label{Thm:WSMedian}
 Under assumptions [A1]-[A3], we have 
 \ban 
 n^{1/2} (\hat{q}_{nW} - q_{0}) & \draro N(0, \Psi_{2W}^{-1} \Psi_{1W} \Psi_{2W}^{-1}), 
 \text{ where } \\
 \Psi_{2W} & = 
 \BE W (X, \BF)\Bigl[  |X - q_0|^{-1} ( \BI_{p} - S (X; q_{0}) S^{T} (X; q_{0}))  \Bigr] 
 \\
 \Psi_{1W}  & = \BE   W^{2} (X, \BF)  S (X; q_{0}) S^{T} (X; q_{0}).
 \ean
 \end{Theorem}

%\rework{
Thus, under very standard regularity conditions, the sample weighted spatial median 
is consistent and is asymptotically normal. Theorem~\ref{Thm:WSMedian}  can be proved in several different ways. Here we use techniques following \cite{ref:AoS891631_Haberman, ref:AoS921514_Niemiro}. Specifically, following Theorem 4 in \cite{ref:AoS921514_Niemiro}, which traces back to \cite{ref:AoS891631_Haberman} with slightly relaxed conditions, we get
%
\begin{align*}
n^{1/2} (\hat{q}_{nW} - q_{0}) =
- \frac{\Psi_{2W}}{\sqrt n} \sum_{i=1}^n W (X_i, \BF)  S (X_i; q_0) + o_P(1).
\end{align*}
%
Theorem~\ref{Thm:WSMedian} follows by applying the central limit theorem.

\paragraph{Remark.} Note that for the result in Theorem~\ref{Thm:WSMedian} to go through, it is not necessary for $\BF$ to be the distribution of $X$. Other choices of $\BF$, e.g. \cite{StatPaper18},  may lead to interesting interpretations of $W(. ,\BF)$ and the resulting location estimator and can be explored further.
%}

@@

It is possible to make the above asymptotic relative efficiency to tend to infinity with 
$p$, for example by choosing $k = \exp\{- p^{2} \}$. Such tending to infinity efficiencies 
are also achievable for fixed $k$,  for example, if $|X - q_{0}|^{2}$ is a Gamma random 
variable with shape parameter $ = 2 + \exp\{- p^{2} \}$. We leave it to future work to 
study such efficiencies more generally.

@@

Obtaining affine invariant weight functions is not challenging. Most functions arising in the context of data depths are affine invariant. Corollary~\ref{Cor:Elliptic_ARE} implies that there is a wide frame of distributions and  choices of weight functions where there is a benefit to considering weighted spatial medians. In fact, the choice of weight functions is even broader: only location invariance is required for Corollary~\ref{Cor:Elliptic_ARE} to hold. We only choose to restrict ourselves here because the subsequent analysis uses weights based on affine invariant data depth functions. Also note that in case \eqref{eqn:CorElliptic_ARE_eqn} does not hold for the original form of a weight function, one can scale all weights by an appropriate constant to reduce the value in the left hand side of \eqref{eqn:CorElliptic_ARE_eqn} to satisfy the condition.

In actual computations, in place of $W (x, \BF)$ we propose using $W (x, \BF_{n})$ where
$ \BF_{n}(z) = (np)^{-1} \sum_{i = 1}^{n} \sum_{j=1}^p
\cI (X_{ij} \leq z_j ), z \in \BR^p
$ is the empirical distribution function. Up to first order asymptotics, the analysis remain unchanged from the above for this modified weight function.

>>>>

Suppose the open unit sphere in $\cX$ is given by 
$\textrm{int} {\cX}_{0; 1} = \{ x \in \cX: | x | < 1 \}$, and let
 $u \in \textrm{int} {\cX}_{0; 1}$. We also fix the set of probability measures $\cM$, 
 and select $\BF \in \cM$. Consider a random element $X \in \cX$, and define the 
 function $\Phi (q; X, u, \BF) = W (X, \BF) \bigl\{ | X - q | 
+ \langle u, X - q \rangle \bigr\}$. 
We define the $(u, \BF)$-th \textit{weighted spatial quantile} of $\cX$ 
as the minimizer $q (u, \BF) \in \cX$ of the expectation of $\Phi (q; X, u, \BF)$, 
that is
\baq
\Psi (q; u, \BF) = \BE \Bigl[ W (X, \BF) \bigl\{ | X - q | 
+ \langle u, X - q \rangle \bigr\} \Bigr] = \BE \Phi (q; X, u, \BF).
\label{eq:WeightedQuantile}
\eaq
%
%\rework{
\noindent
This is a natural generalization of the spatial median \cite{ref:JASA96862_Chaudhuri, ref:Biometrika48414_Haldane,ref:AoS97435_Koltchinskii} ($W(X, \BF) \equiv 1$ and $u = {\bf 0}_p$), or more general spatial quantiles \cite{ref:AoS17591_Cardotetal_Median_HilbertSpace, ref:AoS141203_ChakrabortyChaudhuri_Banach_Quantile, ref:Bernoulli152308_Minsker_Median_Banach} ($W(X, \BF) \equiv 1$). We assume that 
$\Phi (q; X, u, \BF)$ is convex in $q$ for $\BF$-almost all values of $x \in \cX$.

@@

\item[(A3)] $ {\frac{\partial^{2}} {\partial q^{2}}} \Psi (q; 0, \BF)$
exists for all $q$ in a neighborhood of $q_{0}$, and we use the notations
\ban
\Psi_{1W}   & = 
\Bigl( {\frac{\partial} {\partial q}} \Psi (q_{0}; 0, \BF) \Bigr)
 \Bigl( {\frac{\partial} {\partial q}} \Psi (q_{0}; 0, \BF) \Bigr)^{T}, \\
\Psi_{2W}   & = {\frac{\partial^{2}} {\partial q^{2}}} \Psi (q_{0}; 0, \BF). 
\ean
 \end{description}
 These assumptions are very broad-based and general. The first one essentially requires 
 the existence of a population parameter, the second one requires that the minimization 
 approach is meaningful in the population, and the third one essentially requires that the 
 weight function has a finite variance. No further restrictions are placed on the tuning parameter $\BF$ or the choice of the weight function.

 \begin{Theorem}\label{Thm:WSMedian}
 Under assumptions [A1]-[A3], we have 
 \ban 
 n^{1/2} (\hat{q}_{nW} - q_{0}) & \draro N(0, \Psi_{2W}^{-1} \Psi_{1W} \Psi_{2W}^{-1}).
 \ean
 \end{Theorem}

%\rework{
Thus, under very standard regularity conditions, the sample weighted spatial median 
is consistent and is asymptotically normal. Theorem~\ref{Thm:WSMedian}  can be proved in several different ways. Here we use techniques following \cite{ref:AoS891631_Haberman, ref:AoS921514_Niemiro}. Specifically, following Theorem 4 in \cite{ref:AoS921514_Niemiro}, which traces back to \cite{ref:AoS891631_Haberman} with slightly relaxed conditions, we get
%
\begin{align*}
n^{1/2} (\hat{q}_{nW} - q_{0}) =
- \frac{\Psi_{2W}}{\sqrt n} 
\sum_{i=1}^n \nabla \Phi (q; X_{i}, u, \BF)
+ o_P(1),
\end{align*}
%
where $\nabla \Phi (q; X_{i}, u, \BF)$ is any measurable subgradient of 
$\Phi (q; X_{i}, u, \BF)$. 
Theorem~\ref{Thm:WSMedian} follows by applying the central limit theorem.

\paragraph{Remark.} Note that the technical conditions for the result presented in Theorem~\ref{Thm:WSMedian} is one of several alternatives that can conceived, and the scope of this result is broader than what is presented above. First, note that if $\BF$ 
and $\BF_{X}$ are different and the weights are not a function of $q$, 
a situation that may arise in hypothesis testing problems where the weights are based 
on the null distribution, the convexity of $\Phi (q; X_{i}, u, \BF)$ follows automatically 
and is not an assumption. Second, even if $\Phi (q; X_{i}, u, \BF)$ is not convex but sufficiently smooth, we can have a central limit theorem, for example, by using techniques similar to \cite{ref:CBose_AoS05414}. 
Choices of $\BF$ other than $\BF_{X}$, e.g. \cite{StatPaper18},  may lead to interesting interpretations of $W(. ,\BF)$ and the resulting location estimator and will be explored further in future.

@@

%A context where 
It is possible to make the above asymptotic relative efficiency to tend to infinity with 
$p$, for example by choosing $k = \exp\{- p^{2} \}$. Such tending to infinity efficiencies 
are also achievable for fixed $k$,  for example, in the context of some hypothesis tests.
%ing problems.
%f $|X - q_{0}|^{2}$ is a Gamma random 
%variable with shape parameter $ = 2 + \exp\{- p^{2} \}$. 
We leave it to future work to 
study such efficiencies more generally.

@@

Obtaining affine invariant weight functions is not challenging. Most functions arising in the context of data depths are affine invariant. Corollary~\ref{Cor:Elliptic_ARE} implies that there is a wide frame of distributions and  choices of weight functions where there is a benefit to considering weighted spatial medians. In fact, Corollary~\ref{Cor:Elliptic_ARE} can be established under just 
location invariance condition on weights, we use the stronger affine invariance condition
here only because the subsequent sections use weights based on affine invariant 
data depth functions. Also note that in case \eqref{eqn:CorElliptic_ARE_eqn} does not hold for the original form of a weight function, one can scale all weights by an appropriate constant to reduce the value in the left hand side of \eqref{eqn:CorElliptic_ARE_eqn} to satisfy the condition.

In actual computations, in place of $W (x, \BF)$ we propose using $W (x, \BF_{n})$ where
$ \BF_{n}(z) = n^{-1} \sum_{i = 1}^{n} \prod_{j=1}^{p}
\cI (X_{ij} \leq z_j ), z \in \BR^p
$ is the empirical distribution function. Up to first order asymptotics, the analysis remain unchanged from the above for this modified weight function.
